\documentclass[12pt,a4paper]{article}
\setlength{\topmargin}{-5mm}
\setlength{\textheight}{244mm}
\setlength{\oddsidemargin}{0mm}
\setlength{\textwidth}{165mm}

\usepackage{fancyvrb}
\DefineVerbatimEnvironment{verbatim}{Verbatim}{xleftmargin=4mm}

\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{pxfonts}
\usepackage{textcomp}
\usepackage{tabu}

\title{Dijkstra Maps in Roguelikes}
\author{Jeremy Mates}
\date{January 11, 2020}

\usepackage{hyperref}
\hypersetup{pdfauthor={Jeremy Mates},pdftitle={Dijkstra Maps in Roguelikes}}

\begin{document}
\bibliographystyle{plainnat}
\maketitle

\setlength{\parindent}{0pt}

\section*{Simple Pathfinding}

Consider the following map, wherein there are three velociraptors
\texttt{R} surrounding the player \texttt{@} on an open
field\cite{xkcd135}. How can the raptors reach the player?

\begin{verbatim}
......R
.R.....
.......
...@...
.......
.......
..R....
\end{verbatim}

One solution to this problem is the so-called Dijkstra Map\cite{tipodm}.
A map this small can easily be worked out on graph paper; set the
location of the player to the value \texttt{0}. Then in the squares
adjacent to that point write the next highest number, \texttt{1}.

\begin{verbatim}
.......
.......
...1...
..101..
...1...
.......
.......
\end{verbatim}

And so forth until the map is filled. This is a Dijkstra Map.

\begin{verbatim}
6543456
5432345
4321234
3210123
4321234
5432345
6543456
\end{verbatim}

Note that diagonal moves were not considered\textendash only those
points North South East or West of the given square. This has
ramifications on more complicated maps that contain obstacles.
Meanwhile, each velociraptor can use this map to find the player by
walking ``downhill'' to \texttt{0}, or can flee by picking a cell with a
higher value.

\section*{Obstacles}

Many maps will be more complicated than the above and will contain
obstacles, typically walls, or lava, or there may be monsters that can
pass through walls but not across holy water\textendash specters
perhaps\textendash or anything else you can imagine. Let us for now
only consider physical walls, represented using \texttt{\symbol{35}} as
is typical for roguelikes.

\begin{verbatim}
R....
.###.
.#@..
.#.##
.#.#.
...#.
\end{verbatim}

A Dijkstra Map for this might look like

\vskip 1em%
\begin{tabu} spread 0pt{|X|X|X|X|X|} \hline
8 & 7 & 6 & 5 & 4\\\hline
9 & -1 & -1 & -1 & 3\\\hline
8 & -1 & 0 & 1 & 2\\\hline
7 & -1 & 1 & -1 & -1\\\hline
6 & -1 & 2 & -1 & ?\\\hline
5 & 4 & 3 & -1 & ?\\\hline
\end{tabu}
\vskip 1em%

where the impassable walls are represented by \texttt{-1}. One may use
\texttt{Inf} or an object for such cells, but I assume here an integer
or fixnum array. Another option would be to leave walls at the maximum
integer or \texttt{MOST-POSITIVE-FIXNUM} value and to ignore them while
calculating the costs. Note the two squares lower right that have no
path to the player; these after calculation will typically remain at the
maximum integer value the field was filled with before the pathfinding
pass. This map by the way was made with my \texttt{Game::DijkstraMap}
module\cite{gdm}.

\begin{verbatim}
#!/usr/bin/env perl
use strict;
use warnings;
use Game::DijkstraMap;
my $dm = Game::DijkstraMap->new( str2map => <<'EOM' );
.....
.###.
.#x..
.#.##
.#.#.
...#.
EOM
print $dm->to_tsv;
\end{verbatim}

Pathfinding on this map must ignore \texttt{-1} and look for values
equal to or greater than \texttt{0} when routing to a goal. Here, the
raptor should move horizontally to reach the player in as few moves
as possible.

\section*{Multiple Goals}

A Dijkstra Map may contain multiple goals; pathfinding will find the (or
a) nearest goal. Crabs \texttt{c} for example may wish to retreat to
water \texttt{\symbol{126}} when threatened by the player.

\begin{verbatim}
..~~......~.....
.~~~~......~~...
...~~..c..~~....
..........c.....
...~~...........
..~~...c........
...~~...........
.....~.........@
\end{verbatim}

\vskip 1em%
\begin{tabu} spread 0pt{|X|X|X|X|X|X|X|X|X|X|X|X|X|X|X|X|} \hline
2 & 1 & 0 & 0 & 1 & 2 & 3 & 4 & 2 & 1 & 0 & 1 & 1 & 2 & 3 & 4\\\hline
1 & 0 & 0 & 0 & 0 & 1 & 2 & 3 & 4 & 2 & 1 & 0 & 0 & 1 & 2 & 3\\\hline
2 & 1 & 1 & 0 & 0 & 1 & 2 & 3 & 2 & 1 & 0 & 0 & 1 & 2 & 3 & 4\\\hline
3 & 2 & 2 & 1 & 1 & 2 & 3 & 4 & 3 & 2 & 1 & 1 & 2 & 3 & 4 & 5\\\hline
4 & 3 & 1 & 0 & 0 & 1 & 2 & 3 & 4 & 3 & 2 & 2 & 3 & 4 & 5 & 6\\\hline
2 & 1 & 0 & 0 & 1 & 2 & 3 & 4 & 5 & 4 & 3 & 3 & 4 & 5 & 6 & 7\\\hline
3 & 2 & 1 & 0 & 0 & 1 & 2 & 3 & 4 & 5 & 4 & 4 & 5 & 6 & 7 & 8\\\hline
4 & 3 & 2 & 1 & 1 & 0 & 1 & 2 & 3 & 4 & 5 & 5 & 6 & 7 & 8 & 9\\\hline
\end{tabu}
\vskip 1em%

This map however does not consider other crabs\textendash can two occupy
the same square, or will one need to find a longer route to a free water
cell?\textendash nor the influence of the player; a monster may
realistically have multiple desires: get to water while also avoiding
the player. This can be solved in various ways; moves could be ranked by
utility value--corners might be scored poorly for a fleeing monster (see
e.g. ``Behavioral Mathematics for Game AI''\cite{gameai} for ideas)--or
by using a combination of maps.

\section*{Combining Dijkstra Maps}

Multiple maps may be added together and the combined score used to move
an entity towards or away from goals. Suppose that some
Knight\cite{donq} seeks to make justice upon a fleeing goblin. The
monster, however, has been cornered in a room so simple ``move to the
highest cost cell'' code will leave the monster trapped in a corner.
There should be, in this condition, some means for the monster to slip
past the player.

\begin{verbatim}
######
#....#
#.g.@.
#....#
######
\end{verbatim}

A monster whose state is fleeing needs to be anywhere else; this can be
represented by a map with the monster as the goal, higher values thus
being more desirable. This map can be combined with the map centered on
the player--and also any of their allies, if necessary. Since the
player map will have undesirably low values close to the player, the
resulting map should give a path of higher values for a monster to flee
along step by step.

\begin{verbatim}
######   ######   ######
#2123#   #4321#   #6444#
#1g123 + #321@1 = #42224
#2123#   #4321#   #6444#
######   ######   ######
\end{verbatim}

The best single move in this case would be for the goblin to move to
one of the corners (combined cost 6) and for the player to step
towards them, so

\begin{verbatim}
######   ######
#g...#   #....#
#.....   #..@..
#....#   #....#
######   ######
 g123     3212     3335 
 11345 +  21@12 =  33357
 2345     3212     5557 
\end{verbatim}

The goblin is now stuck on an island (cost 3), so the player
advances again,

\begin{verbatim}
######   ######
#g...#   #....#
#.....   #.@...
#....#   #....#
######   ######
 g123     2123     2246 
 12345 +  1@123 =  22468
 2345     2123     4468 
\end{verbatim}

This is not productive as even with combined maps the goblin still gets
stuck. Perhaps if we give more weight to the values of one map or the
other, say multiplying the goblin costs by 2?

\begin{verbatim}
######   ######
#g...#   #....#
#.....   #.@...
#....#   #....#
######   ######
 g123     2123     2369 
 11234 +  1@123 =  347AD
 2223     2123     67AD 
  * 2
\end{verbatim}

This looks promising, though the best move is to move into the player,
which might be bad. The easiest solution would be to treat the player
(and any allies) as impassable squares on the goblin's flee map

\begin{verbatim}
######   ######
#g...#   #....#
#.#...   #.@...
#....#   #....#
######   ######
 g123     2123     2369 
 1#234 +  1@123 =  3#7AD
 2223     2123     67AD 
  * 2
\end{verbatim}

Better. The goblin might here roll unlucky and move down instead of
right towards the exit, but that should create conditions that will
cause them to try to move past the player:

\begin{verbatim}
######   ######
#....#   #....#
#g#...   #.@...
#....#   #....#
######   ######
 1234     2123     458B 
 1#456 +  1@123 =  1#9CF
 1234     2123     458B 
  * 2
\end{verbatim}

However the monster may move back and forth between the back wall and an
edge of the room instead of only moving towards the door if the RNG
dictates that, but at least it is not always stuck, and it may roll well
and escape. \\

Another method would be to construct a line or path to the highest value
on the monster's map instead of only looking for the single best move in
any given turn; this would avoid the possibility of the monster getting
stuck over multiple turns though may be more expensive to calculate and
store. The monster's map also need not comprise the entire field; it may
only need to be slightly larger than the player's field of vision. \\

Different or non-integer weights can be used, or the maps could be
subtracted instead of being added, though experimentation shows that
adding the maps and applying a small positive integer weight to the
monster flee map better avoids problems such as the monster islanding
itself in a corner. \\

Play testing with different approaches and map weights will likely be
necessary. Level design may also be a factor, 3x3 rooms give an obvious
path for the player to try to box a monster in with; larger rooms or
rooms with multiple exits or rooms with rounded corners should create
different weights for the monsters to flee along.

\section*{The Diagonal}

Costs in the above maps have been done without consideration for
diagonal moves. Consider the following map, where \texttt{x} is the
goal.

\begin{verbatim}
@#..
#...
...x
\end{verbatim}

\vskip 1em%
\begin{tabu} spread 0pt{|X|X|X|X|} \hline
? & -1 & 3 & 2\\\hline
-1 & 3 & 2 & 1\\\hline
3 & 2 & 1 & 0\\\hline
\end{tabu}
\vskip 1em%

The player here cannot pathfind to the goal as the diagonal move was not
considered by the 4-way algorithm that only consults cells North South
East and West. Various roguelikes (Angband, Dungeon Crawl Stone Soup)
permit such diagonal moves, so will need to use an 8-way algorithm when
constructing a Dijkstra Map. Other roguelikes (Brogue, POWDER) may deny
such diagonal moves so can use the 4-way algorithm. This choice also
influences level design. 4-way and 8-way roguelikes require rather
different diagonal corridors:

\begin{verbatim}
8-way    @####   4-way    @.###
corridor #.###   corridor #..##
         ##x##            ##x##
\end{verbatim}

Brogue (like rogue) is complicated in that it allows some diagonal
moves. A 4-way Dijkstra Map algorithm can be used with 8-way motion
provided 4-way moves are possible to everywhere that must be reached.
Diagonal moves in such a case exist as shortcuts\textendash moving
diagonally along the above 4-way corridor (which Brogue does not permit,
nor POWDER unless one is ploymorphed into a grid bug).

\section*{Diagonal Maps}

8-way maps typically assign equal costs to all adjacent squares. The
original raptor map thus looks like:

\begin{verbatim}
......R  3333333
.R.....  3222223
.......  3211123
...@...  3210123
.......  3211123
.......  3222223
..R....  3333333
\end{verbatim}

This while traditional for roguelikes is not actually correct; diagonals
under euclidean geometry should instead use $\sqrt{x^2 + y^2}$ or
$\sqrt{2}$ instead of \texttt{1} for the closest diagonal. Various
roguelikes are actually non-euclidean: Brogue and Dungeon Crawl Stone
Soup apply the same cost to a move in any direction, diagonal or
otherwise. Anyways! Our original diagonal map that stumped the 4-way map
under (non-euclidean) 8-way is:

\begin{verbatim}
@#..
#...
...x
\end{verbatim}

\vskip 1em%
\begin{tabu} spread 0pt{|X|X|X|X|} \hline
3 & -1 & 2 & 2\\\hline
-1 & 2 & 1 & 1\\\hline
3 & 2 & 1 & 0\\\hline
\end{tabu}
\vskip 1em%

And the player can path to the goal.

\section*{Not Just for Animates}

Dijkstra Map can be used for other purposes. Given a map with two rooms
in it, how to connect them?

\begin{verbatim}
####################
###########XXXXXX###
###########X....X###
###########X....X###
#XXXXXX####XXXXXX###
#X....X#############
#X....X#############
#X....X#############
#XXXXXX#############
####################
\end{verbatim}

Place a goal anywhere in the wall space, and then considering only walls
\texttt{\symbol{35}} as passable construct a Dijkstra Map with the goal
as the destination, here shown partially complete.

\begin{verbatim}
           XXXXXX   
           X....X8  
     DCBA98X....X7  
 XXXXXXA987XXXXXX6  
 X....X98765432345  
 X....X87654321234  
 X....X76543210123  
 XXXXXX 765432123   
\end{verbatim}

Then pick a random door location in each room, and pathfind from each
door to the goal, drawing a corridor along the way.

\begin{verbatim}
####################
###########XXXXXX###
###########X....+.##
#####...###X....X.##
#XXXX+X.###XXXXXX.##
#X....X...#######.##
#X....X##.........##
#X....X#######..####
#XXXXXX#############
####################
\end{verbatim}

This will run into complications if rooms are adjacent or worse two
rooms block a third from reaching a goal (there should likely be one
goal per unconnected wall space in a map) though there are various
solutions to these challenges, such as pathfinding through rooms or
using additional code to find and link up adjacent rooms.

\section*{Super Dimensional Dijkstra Maps}

We need not confine ourselves to two, or even three dimensions;
Dijkstra Maps can be built in arbitrary numbers of dimensions
(memory requirements, implementation demands, and sanity
permitting). The following is a four-dimensional map with an
implementation\cite{crazylisp} that does not consider diagonal
moves legal.

\begin{verbatim}
% sbcl --noinform --load dijkstramap.lisp
* (setf *dimap-cost-max* 99)

99
* (defparameter level
    (make-array '(3 3 3 3)
                 :initial-contents
                 '((((99 -1 99) (-1 99 -1) (99 -1 99))
                    ((-1 99 99) (99 99 99) (99 99 -1))
                    ((99 99 99) (99 99 99) (99 99 99)))
                   (((-1 99 99) (99 99 99) (99 99 -1))
                    ((99 99 99) (99  0 99) (99 99 99))
                    ((99 99 99) (99 99 99) (99 99 99)))
                   (((99 99 99) (99 99 99) (99 99 99))
                    ((99 99 99) (99 99 99) (99 99 99))
                    ((99 99 99) (99 99 99) (99 99 99))))))

LEVEL
* (dimap-calc level)

4
* level

#4A((((99 -1 4) (-1 2 -1) (4 -1 99))
     ((-1 2 3) (2 1 2) (3 2 -1))
     ((4 3 4) (3 2 3) (4 3 4)))
    (((-1 2 3) (2 1 2) (3 2 -1))
     ((2 1 2) (1 0 1) (2 1 2))
     ((3 2 3) (2 1 2) (3 2 3)))
    (((4 3 4) (3 2 3) (4 3 4))
     ((3 2 3) (2 1 2) (3 2 3))
     ((4 3 4) (3 2 3) (4 3 4))))
*
\end{verbatim}

Productive uses for such maps are left as an exercise to the reader.

\clearpage
\bibliography{references}
\end{document}
